---
title: "Introduction to the earthquakedata package"
author: "Paul Visser"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to the earthquakedata package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7
)
library(earthquakedata)
library(dplyr)
library(ggplot2)

theme_timeline <- ggplot2::theme_classic() +
  ggplot2::theme(
    # hide the y axis line, ticks and title
    axis.line.y = ggplot2::element_blank(),
    axis.ticks.y = ggplot2::element_blank(),
    axis.title.y = ggplot2::element_blank(),
    # change the y axis line text to be darkgray, size 11 with a right-hand margin of 15
    axis.text.y = ggplot2::element_text(colour = "darkgray",size = 11, margin = ggplot2::margin(r = 15)),
    # show x axis and make the line colour black with a thickness of 1
    axis.line.x = ggplot2::element_line(colour = "black", size = ggplot2::rel(1)),
    # change the x axis text to be darkgray, size 11 with a top margin of 3
    axis.text.x = ggplot2::element_text(colour = "darkgray",size = 11, margin = ggplot2::margin(t = 3)),
    # change the x axis title to be size 14 with a top margin of 5
    axis.title.x = ggplot2::element_text(size = 14, margin = ggplot2::margin(t = 5)),
    # as an extra, give the plot a title and justify it horisontally in the middle of the plot
    plot.title = ggplot2::element_text(color = "darkgray", size = 24, hjust = 0.5),
    # position the legend at the bottom of the screen
    legend.position = "bottom",

    complete = TRUE
  )
```

## Aim of the package

The earthquakedata package allows the user to clean and visualise the earthquake data set provided by the U.S. National Oceanographic and Atmospheric Administration (NOAA). This data set contains information about 5,933 earthquakes over an approximately 4,000 year time span.

``` {r, message=FALSE}
readr::read_delim("earthquakes.tsv.gz",delim = "\t") %>%
  eq_clean_data() %>%
  dplyr::filter(COUNTRY == c("USA") & lubridate::year(DATE) >= 2000) %>%
  ggplot() +
    geom_timeline(aes(x = DATE, y = COUNTRY,size = EQ_PRIMARY, colour = DEATHS)) +
    geom_timeline_label(aes(x = DATE, y = COUNTRY, label = LOCATION_NAME, size = EQ_PRIMARY), n_max = 5) +
    ggtitle("Earthquake Timeline") +
    theme_timeline +
    labs(size = "Richter Scale value:", colour = "# of Deaths:")
``` 

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
